The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-07-17 script/de/chapters/paper_and_pencil.tex
added figure to paper_and_pencil appendix
paper and pencil: converted pseudo appendix in to section; defined sagecode float (en only) for sage code examples
CT script, paper and pencil: header updated; defined sagecode float (de only) for sage code examples and added list at the end; improved the code of the first 5 Sage samples (en only yet)

\chapter{Papier- und Bleistift-Verschlüsselungsverfahren}
\label{Kapitel_PaperandPencil}
(Christine Stötzel, April 2004; Updates B.+C. Esslinger, Juni 2005)
\index{Papier- und Bleistiftverfahren}

@@

% ---------------------------------------------------------------------------
\newpage
\section*{Appendix A: Examples using Sage}
\addcontentsline{toc}{section}{Appendix A: Examples using Sage}
\index{Sage}

\subsection*{Transposition ciphers using Sage}

>>>>

\chapter{Papier- und Bleistift-Verschlüsselungsverfahren}
\label{Kapitel_PaperandPencil}
(Christine Stötzel, April 2004; Updates B.+C. Esslinger, Juni 2005; Updates B. Esslinger, Juli 2009)
\index{Papier- und Bleistiftverfahren}

@@

% ---------------------------------------------------------------------------
\newpage
\section{Beispiele mit Sage}
\index{Sage}

Im Folgenden werden einige der klassischen Verfahren mit dem Computer-Algebra-System Sage implementiert. Der Code ist lauffähig und getestet mit Sage Version 4.1.

Eine erste Einführung in das CAS Sage finden Sie im Anhang \ref{s:appendix-using-sage}.

Um die Beispielprogramme leichter lesbar zu machen, hielten wir uns an die in der folgenden
Grafik aufgeführten Bezeichnungen und Abläufe:
\begin{itemize}
  \item Die Verschlüsselung besteht aus den Schritten codieren und verschlüsseln.
  \item Beim Codieren werden in dem gegebenen Klartext P die Zeichen auf Groß- bzw.
        Kleinbuchstaben angepasst (je nach gegebenem Alphabet), und alle
	Nichtalphabetzeichen werden entfernt.
  \item Durch die Verschlüsselung wird das Chiffrat C erzeugt.
  \item Die Entschlüsselung besteht ebenfalls aus den zwei Schritten entschlüsseln und decodieren.
  \item Decodieren ist nur nötig, wenn die benutzten Symbole im Alphabet nicht die ASCII-Zeichen
        sind ????????????????.
\end{itemize}

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{figures/encryption-decryption-de}
\caption{Struktur und Namenskonventionen in den Sage Programmbeispielen} 
\label{XXX}
\end{center}
\end{figure}

\subsection{Transposition ciphers using Sage}

### 2009-07-17 script/en/chapters/paper_and_pencil.tex
CT script, paper and pencil: header updated; defined sagecode float (de only) for sage code examples and added list at the end; improved the code of the first 5 Sage samples (en only yet)

\chapter{Paper and Pencil Encryption Methods}
\label{Kapitel_PaperandPencil}
(Christine St\"otzel, April 2004; Updates: B.+C. Esslinger, June 2005)
\index{Paper- and pencil methods}

@@

@@

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{figures/encryption-decryption-en}
\caption{XXX} 
\label{XXX}
\end{center}
\end{figure}

Tranposition ciphers are implemented in the class
\begin{center}
\verb!sage.crypto.classical.TranspositionCryptosystem!
\end{center}
To construct and work with a transposition cipher, we first need to
determine the alphabet that contains our plaintext and ciphertext. 
Typically, this alphabet will be the upper-case letters of the
English alphabet, which can be accessed via the function
\begin{center}
\verb!sage.monoids.string_monoid.AlphabeticStrings!
\end{center}
We then need to decide on the block length of a block permutation,
which is the length of the row vector to be used in encryption and
decryption. This row vector is our key, and it specifies a permutation
of a plaintext.

Here is a transposition cipher of block length 14, where every letter
in the plaintext is shifted to the left by one character, with wrap
around. That is the encryption process. The decryption process is
shifting to the left by $14 - 1 = 13$ characters in order to recover
the plaintext.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher of block length 14
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # our encryption/decryption key
sage: key = [2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1]
sage:
sage: # encrypt a plaintext
sage: E = T(key)
sage: C = E(T.encoding("a b c d e f g h i j k l m n")); C
BCDEFGHIJKLMNA
sage:
sage: # decrypt ciphertext
sage: keyInv = [14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13]
sage: D = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
\end{Verbatim}
\caption{Caption 1}
\end{sagecode}

The encryption key can also be such that it shifts each letter of the
plaintext to the left by two characters, with wrap around. Then we
need to shift each letter of the ciphertext to the left by $14 - 2 =
12$ characters in order to obtain the plaintext.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # encrypt by shifting to the left by 2 letters
sage: key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage: E = T(key)
sage: C = E(T.encoding("a b c d e f g h i j k l m n")); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt by shifting to the left by 12 letters
sage: keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: D = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
\end{Verbatim}
\caption{Caption 2}
\end{sagecode}

@@

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage:
sage: # encryption
sage: key = T.random_key(); key
(1,2,7)(3,9)(4,5,10,12,8,13,11)(6,14)
sage: P = T.encoding("a b c d e f g h i j k l m n")
sage: C = T.enciphering(key, P); C
BGIEJNAMCLDHKF
sage:
sage: # decryption
sage: T.deciphering(key, C)
ABCDEFGHIJKLMN
sage: T.deciphering(key, T.enciphering(key, P))
ABCDEFGHIJKLMN
\end{Verbatim}

@@

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # the plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage:
sage: # substitution cipher with random key
sage: S = SubstitutionCryptosystem(A)
sage: key = S.random_key(); key
CMPGOWFIJRESXNYBKUDALQHTVZ
sage:
sage: # encryption
sage: msg = A.encoding("Substitute this with something else better."); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: S.enciphering(key, msg)
DLMDAJALAOAIJDHJAIDYXOAIJNFOSDOMOAAOU
\end{Verbatim}

We now decrypt the ciphertext to recover our plaintext.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # decryption
sage: plaintext = S.deciphering(key, ciphertext); plaintext
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: plaintext == msg
True
\end{Verbatim}

@@

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....: 20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: encrypt = S(key)
sage: msg = A.encoding("Shift three positions to the right."); msg
SHIFTTHREEPOSITIONSTOTHERIGHT
sage: ciphertext = encrypt(msg); ciphertext
VKLIWWKUHHSRVLWLRQVWRWKHULJKW
\end{Verbatim}

Next, we recover the plaintext.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....: 14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv)
sage: plaintext = decrypt(ciphertext); plaintext
SHIFTTHREEPOSITIONSTOTHERIGHT
sage: plaintext == msg
True
\end{Verbatim}

@@

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # the plaintext/ciphertext alphabet
sage: A = BinaryStrings()
sage: # substitution cipher over the alphabet A
sage: S = SubstitutionCryptosystem(A); S
Substitution cryptosystem on Free binary string monoid
sage: key = S.random_key(); key
10
sage:
sage: # encryption
sage: msg = A.encoding("Working with the binary numbers."); msg
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001110100011010000110010100100000011000100110100101\
10111001100001011100100111100100100000011011100111010101101101011000100110010\
1011100100111001100101110
sage: ciphertext = S.enciphering(key, msg); ciphertext
10101000100100001000110110010100100101101001000110011000110111111000100010010\
11010001011100101111101111110001011100101111001101011011111100111011001011010\
01000110011110100011011000011011011111100100011000101010010010100111011001101\
0100011011000110011010001
sage:
sage: # decryption
sage: plaintext = S.deciphering(key, ciphertext); plaintext
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001110100011010000110010100100000011000100110100101\
10111001100001011100100111100100100000011011100111010101101101011000100110010\
1011100100111001100101110
sage: plaintext == msg
True
\end{Verbatim}

@@

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: A = HexadecimalStrings()
sage: S = SubstitutionCryptosystem(A)
sage: key = S.random_key(); key
26849de70f35b1ca
sage: msg = A.encoding("Working with a larger alphabet."); msg
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: ciphertext = S.enciphering(key, msg); ciphertext
d7ea78e5efece78277ef79e082e682ebe678e7ed7882e6eb72e0e6e8ed798c
sage: plaintext = S.deciphering(key, ciphertext); plaintext
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: plaintext == msg
True
\end{Verbatim}

@@

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct Vigenere cipher
sage: A = AlphabeticStrings()
sage: V = VigenereCryptosystem(A, 19); V
Vigenere cryptosystem on Free alphabetic string monoid on A-Z of
period 19
sage: key = V.random_key(); key
ETKAAYJBKURHAUJXOMC
sage:
sage: # encryption
sage: msg = V.encoding("The Vigenere cipher is polyalphabetic."); msg
THEVIGENERECIPHERISPOLYALPHABETIC
sage: ciphertext = V.enciphering(key, msg); ciphertext
XAOVIENOOLVJIJQBFUUTHVYAJYIKVVAIW
sage:
sage: # decryption
sage: plaintext = V.deciphering(key, ciphertext); plaintext
THEVIGENERECIPHERISPOLYALPHABETIC
sage: plaintext == msg
True
\end{Verbatim}

@@

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # construct a Hill cipher
sage: A = AlphabeticStrings()
sage: H = HillCryptosystem(A, 19); H
Hill cryptosystem on Free alphabetic string monoid on A-Z of block
length 19
sage: key = H.random_key(); key

[ 6 25  1 12  6 20 22 25  8 22 14 19 19  5  9 21 22  1 22]
[ 3  2  7 17 14 17  5 11 20  6  5  1 14 14 21 18  1 20  4]
[ 6 15 24  7  4 23 23 16 12  3 11 20 21  2  2 10 16  4 17]
[ 8 19  9 19 20  6 12  1 17 11 25 24 14 10  2 15  4  3 25]
[25 24 15 12  6  7 12 10  7 17 16 11 11 18 13 24 18 25 20]
[16 11 11 12  3  6 14 25 24  3  4 14 13 14 21  1  5  1 24]
[25  9  0 16  2 22  8  3  1  4 12 22 22 14  0  3 20  2 15]
[ 7 19 17 25 10 22  8  9  0  5 25  2 17 23  3 18 18  7  9]
[15  7  8 12  0 12 24 23 18 21  9 24 24 20 13 13  8 24  9]
[11 21 11 25  0  2 23 25  2  8  1  9 20  7 14  1 24  4 10]
[21  8 14  8 12 24 14  2 19  1  7 19  1 10 24  5  9  1  9]
[ 4  5 23 24  3  4 17 22  3 12  1 12  0 21 12  3 12 18  6]
[22 25 16 15 14  1  1 23  7 24  8 14 14 18 17 23 20 23 22]
[11  6 16  5  3  6 14  8 13 17 12 21  3 16  0  7  7 14 14]
[18 21  5 14 13 12  7  9  2 17 11 20 25 12  7 18 19 20  1]
[18 20 19  9  8  7 18 21 15 17 17 20 15 11  0  0  4 18 24]
[ 3  4 18  9 18  4  9 21 23 25  0  2 22 16  9  0 23 20 12]
[20 21 14  6  0  6 15 19 20 19 17  1 18 12  5 11  0  2 10]
[ 3  0 13 18 24 22 19  1 19 18  8 23  3 13 15 16  6 17  3]
sage:
sage: # encryption
sage: msg = H.encoding("Hill or matrix cipher uses matrix operations.")
sage: ciphertext = H.enciphering(key, msg); ciphertext
GJLILYPELTCHRKNXSTGJYCZSVNMPTIUZCTOJPU
sage:
sage: # decryption
sage: plaintext = H.deciphering(key, ciphertext); plaintext
HILLORMATRIXCIPHERUSESMATRIXOPERATIONS
sage: plaintext == msg
True
\end{Verbatim}

>>>>

\chapter{Paper and Pencil Encryption Methods}
\label{Kapitel_PaperandPencil}
(Christine St\"otzel, April 2004; Updates: B.+C. Esslinger, June 2005; Updates B. Esslinger, July 2009)
\index{Paper- and pencil methods}

@@

In the following chapter some classic ciphers are implemented using the computer algebra system
Sage. The code was tested with Sage vrsion 4.1.

A first introduction to the CAS Sage can be found in the appendix \ref{s:appendix-using-sage}.

To make the sample code easier to understand, we used the structure and the naming conventions
shown in the graphics below:
\begin{itemize}
  \item Encryption consists of the two steps encoding and enciphering.
  \item Encoding adapts the letters in the given plaintext P to the case defined in the
        given alphabet, and all non-alphabet characters are filtered out.
  \item Enciphering creates the ciphertext C.
  \item Decryption also consists of two steps: deciphering and decoding.
  \item Decoding is only necessary if the symbols in the alphabet are no ASCII characters ???????????????? xxxxxxxxxxxxx.
\end{itemize}

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{figures/encryption-decryption-en}
\caption{Structure and naming convention of the Sage code examples} 
\label{XXX}
\end{center}
\end{figure}

@@

Tranposition ciphers are implemented in the class
\begin{center}
\verb!sage.crypto.classical.TranspositionCryptosystem!
\end{center}
To construct and work with a transposition cipher, we first need to
determine the alphabet that contains the symbols used to build our plaintext and ciphertext. 
Typically, this alphabet will be the upper-case letters of the
English alphabet, which can be accessed via the function
\begin{center}
\verb!sage.monoids.string_monoid.AlphabeticStrings!
\end{center}
We then need to decide on the block length of a block permutation,
which is the length of the row vector to be used in encryption and
decryption. This row vector is our key, and it specifies a permutation
of a plaintext.

Here is a transposition cipher of block length 14, where every letter
in the plaintext is shifted to the left by two characters, with wrap
around. That is the encryption process. The decryption process is
shifting each letter of the ciphertext to the left by $14 - 2 = 12$
characters in order to recover the plaintext.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14
sage: T = TranspositionCryptosystem(AlphabeticStrings(), 14)
sage: # given plaintext
sage: P   = "a b c d e f g h i j k l m n"
sage: # encryption key
sage: key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars, convert lower-case into upper-case)
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in two steps)
sage: E   = T(key)
sage: C   = E(msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: D   = T(keyInv)
sage: D(C)
ABCDEFGHIJKLMN
sage:
sage: # Representation of key and inverse key as permutations
sage: E
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: D
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple Transposition by shifting (key and inverse key explicitely given)}
\end{sagecode}

The next sample is also a simple column transposition. But now the code is a little bit more automated in generating the keys from the shift parameter.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # transposition cipher using a block length of 14, code more variable
sage: keylen = 14
sage: shift = 2
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen)
sage:
sage: # construct the plaintext string from the first 14 letters of the alphabet plus blanks
sage: # plaintext   = "A B C D E F G H I J K L M N"
sage: A.gens()
(A, B, C, D, E, F, G, H, I, J, K, L, M, N, O, P, Q, R, S, T, U, V, W, X, Y, Z)
sage: P=''
sage: for i in range(keylen): P=P + " " + str(A.gen(i))
....:
sage: P
' A B C D E F G H I J K L M N'
sage:
sage: # encryption key
sage: # key = [3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage: key = [(i+shift).mod(keylen) + 1 for i in range(keylen)]; key
[3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1, 2]
sage:
sage: # encode plaintext (get rid of non-alphabet chars)
sage: # Rem: Encoding is not necessary, if P only contains capital letters
sage: msg = T.encoding(P)
sage: # encrypt plaintext by shifting to the left by 2 letters (do it in one step)
sage: C   = T.enciphering(key, msg); C
CDEFGHIJKLMNAB
sage:
sage: # decrypt ciphertext by shifting to the left by 12 letters
sage: # keyInv = [13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) + 1 for i in range(keylen)]; keyInv
[13, 14, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12]
sage: DC   = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMN
sage:
sage: # using the deciphering method requires to change the type of the variable key
sage: DC  = T.deciphering(T(key).key(), C); DC
ABCDEFGHIJKLMN
sage:
sage: # representation of key and inverse key as permutations
sage: T(key)
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(key).key()
(1,3,5,7,9,11,13)(2,4,6,8,10,12,14)
sage: T(keyInv)
(1,13,11,9,7,5,3)(2,14,12,10,8,6,4)
\end{Verbatim}
\caption{Simple Transposition by shifting (key and inverse key constructed with "range")}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # Remark: Enciphering here requires, that the length of msg is a multiple of keylen
sage: keylen = 14   # length of key
sage: A = AlphabeticStrings()
sage: T = TranspositionCryptosystem(A, keylen); T
Transposition cryptosystem on Free alphabetic string monoid on A-Z of block length 14
sage:
sage: P = "a b c d e f g h i j k l m n o p q r s t u v w x y z a b"
sage: key = T.random_key(); key
(1,2,3,13,6,5,4,12,7)(11,14)
sage: msg = T.encoding(P); msg
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage: C   = T.enciphering(key, msg); C
BCMLDEAHIJNGFKPQAZRSOVWXBUTY
sage: DC  = T.deciphering(key, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Just another way of decryption: Using "enciphering" with the inverse key
sage: keyInv = T.inverse_key(key); keyInv
(1,7,12,4,5,6,13,3,2)(11,14)
sage: DC     = T.enciphering(keyInv, C); DC
ABCDEFGHIJKLMNOPQRSTUVWXYZAB
sage:
sage: # Test correctness of decryption
sage: msg == DC
True
\end{Verbatim}
\caption{Simple Column Transposition with randomly generated (permutation) key}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A   = AlphabeticStrings()
sage: S   = SubstitutionCryptosystem(A)
sage:
sage: P   = "Substitute this with something else better."
sage: key = S.random_key(); key
INZDHFUXJPATQOYLKSWGVECMRB
sage:
sage: # method encoding can be called from A or from T
sage: msg = A.encoding(P); msg
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: C   = S.enciphering(key, msg); C
WVNWGJGVGHGXJWCJGXWYQHGXJOUHTWHNHGGHS
sage:
sage: #.# We now decrypt the ciphertext to recover our plaintext.
sage:
sage: DC  = S.deciphering(key, C); DC
SUBSTITUTETHISWITHSOMETHINGELSEBETTER
sage: msg == DC
True
\end{Verbatim}
\caption{Monoalphabetic Substitution with randomly generated key}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
....:          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: encrypt = S(key); encrypt
DEFGHIJKLMNOPQRSTUVWXYZABC
sage: C       = encrypt(msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #.# Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
....:             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: decrypt = S(keyInv); decrypt
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC      = decrypt(C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; key explicitely given, step-by-step approach)}
\end{sagecode}

The next sample does the same, but the code is more sophisticated/automated.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # plaintext/ciphertext alphabet
sage: A = AlphabeticStrings()
sage: keylen = len(A.gens()); keylen
26
sage: shift  = 3
sage: P = "Shift the alphabet three positions to the right."
sage:
sage: # construct Caesar cipher
sage: S = SubstitutionCryptosystem(A)
sage: ## IDEE: key mit Range (26).shift(3) konstruieren statt wie in Sample02
sage: ## Shift und Map gabe es nicht als interne Methode, also wie in Sample 2
sage: # key = A([3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, \
sage: #          20, 21, 22, 23, 24, 25, 0, 1, 2])
sage: key = [(i+shift).mod(keylen) for i in range(keylen)];
sage: key = A(key); key
DEFGHIJKLMNOPQRSTUVWXYZABC
sage:
sage: # encrypt message
sage: msg     = A.encoding(P); msg
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage: C       = S.enciphering(key, msg); C
VKLIWWKHDOSKDEHWWKUHHSRVLWLRQVWRWKHULJKW
sage:
sage: #.# Next, we recover the plaintext.
sage:
sage: # decrypt message
sage: # keyInv = A([23, 24, 25, 0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, \
sage: #             14, 15, 16, 17, 18, 19, 20, 21, 22])
sage: shiftInv=keylen-shift;
sage: keyInv = [(i+shiftInv).mod(keylen) for i in range(keylen)];
sage: keyInv = A(keyInv); keyInv
XYZABCDEFGHIJKLMNOPQRSTUVW
sage: DC     = S.enciphering(keyInv, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: # Just another way of decryption: Using "deciphering" with the key
sage: DC     = S.deciphering(key, C); DC
SHIFTTHEALPHABETTHREEPOSITIONSTOTHERIGHT
sage:
sage: msg == DC
True
\end{Verbatim}
\caption{Caesar (substitution by shifting the alphabet; keys generated)}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: # the plaintext/ciphertext alphabet   xxxxxxxxxxxxxxxxxxxxxxxxxxxxx
sage: A = BinaryStrings()
sage: # substitution cipher over the alphabet A
sage: S = SubstitutionCryptosystem(A); S
Substitution cryptosystem on Free binary string monoid
sage: key = S.random_key(); key
10
sage:
sage: # encryption
sage: msg = A.encoding("Working with the binary numbers."); msg
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001110100011010000110010100100000011000100110100101\
10111001100001011100100111100100100000011011100111010101101101011000100110010\
1011100100111001100101110
sage: ciphertext = S.enciphering(key, msg); ciphertext
10101000100100001000110110010100100101101001000110011000110111111000100010010\
11010001011100101111101111110001011100101111001101011011111100111011001011010\
01000110011110100011011000011011011111100100011000101010010010100111011001101\
0100011011000110011010001
sage:
sage: # decryption
sage: plaintext = S.deciphering(key, ciphertext); plaintext
01010111011011110111001001101011011010010110111001100111001000000111011101101\
00101110100011010000010000001110100011010000110010100100000011000100110100101\
10111001100001011100100111100100100000011011100111010101101101011000100110010\
1011100100111001100101110
sage: plaintext == msg
True
\end{Verbatim}
\caption{xxxxx}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
xxxxxxxxxxxxxxxxxxxxxxxxxx
sage: A = HexadecimalStrings()
sage: S = SubstitutionCryptosystem(A)
sage: key = S.random_key(); key
26849de70f35b1ca
sage: msg = A.encoding("Working with a larger alphabet."); msg
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: ciphertext = S.enciphering(key, msg); ciphertext
d7ea78e5efece78277ef79e082e682ebe678e7ed7882e6eb72e0e6e8ed798c
sage: plaintext = S.deciphering(key, ciphertext); plaintext
576f726b696e6720776974682061206c617267657220616c7068616265742e
sage: plaintext == msg
True
\end{Verbatim}
\caption{xxxxx}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx
sage: # construct Vigenere cipher
sage: A = AlphabeticStrings()
sage: V = VigenereCryptosystem(A, 19); V
Vigenere cryptosystem on Free alphabetic string monoid on A-Z of
period 19
sage: key = V.random_key(); key
ETKAAYJBKURHAUJXOMC
sage:
sage: # encryption
sage: msg = V.encoding("The Vigenere cipher is polyalphabetic."); msg
THEVIGENERECIPHERISPOLYALPHABETIC
sage: ciphertext = V.enciphering(key, msg); ciphertext
XAOVIENOOLVJIJQBFUUTHVYAJYIKVVAIW
sage:
sage: # decryption
sage: plaintext = V.deciphering(key, ciphertext); plaintext
THEVIGENERECIPHERISPOLYALPHABETIC
sage: plaintext == msg
True
\end{Verbatim}
\caption{xxxxx}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
xxxxxxxxxxxxxxxxxxxxxx
sage: # construct a Hill cipher
sage: A = AlphabeticStrings()
sage: H = HillCryptosystem(A, 19); H
Hill cryptosystem on Free alphabetic string monoid on A-Z of block
length 19
sage: key = H.random_key(); key

[ 6 25  1 12  6 20 22 25  8 22 14 19 19  5  9 21 22  1 22]
[ 3  2  7 17 14 17  5 11 20  6  5  1 14 14 21 18  1 20  4]
[ 6 15 24  7  4 23 23 16 12  3 11 20 21  2  2 10 16  4 17]
[ 8 19  9 19 20  6 12  1 17 11 25 24 14 10  2 15  4  3 25]
[25 24 15 12  6  7 12 10  7 17 16 11 11 18 13 24 18 25 20]
[16 11 11 12  3  6 14 25 24  3  4 14 13 14 21  1  5  1 24]
[25  9  0 16  2 22  8  3  1  4 12 22 22 14  0  3 20  2 15]
[ 7 19 17 25 10 22  8  9  0  5 25  2 17 23  3 18 18  7  9]
[15  7  8 12  0 12 24 23 18 21  9 24 24 20 13 13  8 24  9]
[11 21 11 25  0  2 23 25  2  8  1  9 20  7 14  1 24  4 10]
[21  8 14  8 12 24 14  2 19  1  7 19  1 10 24  5  9  1  9]
[ 4  5 23 24  3  4 17 22  3 12  1 12  0 21 12  3 12 18  6]
[22 25 16 15 14  1  1 23  7 24  8 14 14 18 17 23 20 23 22]
[11  6 16  5  3  6 14  8 13 17 12 21  3 16  0  7  7 14 14]
[18 21  5 14 13 12  7  9  2 17 11 20 25 12  7 18 19 20  1]
[18 20 19  9  8  7 18 21 15 17 17 20 15 11  0  0  4 18 24]
[ 3  4 18  9 18  4  9 21 23 25  0  2 22 16  9  0 23 20 12]
[20 21 14  6  0  6 15 19 20 19 17  1 18 12  5 11  0  2 10]
[ 3  0 13 18 24 22 19  1 19 18  8 23  3 13 15 16  6 17  3]
sage:
sage: # encryption
sage: msg = H.encoding("Hill or matrix cipher uses matrix operations.")
sage: ciphertext = H.enciphering(key, msg); ciphertext
GJLILYPELTCHRKNXSTGJYCZSVNMPTIUZCTOJPU
sage:
sage: # decryption
sage: plaintext = H.deciphering(key, ciphertext); plaintext
HILLORMATRIXCIPHERUSESMATRIXOPERATIONS
sage: plaintext == msg
True
\end{Verbatim}
\caption{xxxxx}
\end{sagecode}
